---
title: "Scrap Work for Credit Risk Logistic Hazard Model"
output: pdf_document
date: "2026-01-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




# Variable Selection

Given the context of the PD model and the situations where such a model is typically deployed (loan acceptance/rejection, loan pricing, expected loss calculations, etc.), we are more concerned with out-of-sample performance than raw statistical significance on the training data. For this reason, we use cross-validation (CV) for model selection instead of deviance tests on the entire dataset.

We still care about the final interpretation of model parameters, so we will also qualitatively consider the interpretability of each model when reviewing the results of CV.

We use rolling CV to compare alternative model specifications for loan age. 
Due to the panel structure of the data, we can't easily use a pre-made function for time series CV such as `timetk::time_series_cv()`, so we implement a custom approach\footnote{Even in Python, we'd have to resort to a similar manual approach.}. Manually implementing cross-validation would be very computationally expensive given our large dataset (6.1M+ rows), so we: 
\begin{enumerate}
  \item make use of the `data.table` package\footnote{The `data.table` package saves a lot of computational cost in large part by avoiding the creation of as many deep copies of dataframes as done in base R or `dplyr`.} to speed up dataframe operations, and
  \item use a single relatively small sample of loans in all folds of CV.
\end{enumerate} 

In particular, we use 10-fold CV with 30 month training windows and 8 month validation windows.

Given the binary outcome of our data, we use binary entropy\footnote{a.k.a. the negative log-likelihood (NLL) for an IID Bernoulli sample, a.k.a. the log loss (as called in `sklearn` in Python).} as our loss function to compare model specifications.\footnote{Isn't this flawed because that loss function assumes data is IID (we have panel data so the same loans are auto-correlated across time)? CHECK THIS}

We proceed with a top-down approach for variable selection.
\begin{itemize}
  \item Loan age is known to typically be a key risk factor, often with nonlinear association, so we use CV for selecting the functional form for the loan age term of the model as described above. 
    \item Moreover, we will actually use Ridge GLM (L2 regularized) instead of the plain GLM model for these CV folds. This is because Ridge offers a solution to these two critical issues: 
      \item Firstly, there will likely be some categorical predictors with only one unique level for some folds, in which case `glm()` will fail to fit the model because it doesn't know how to assign a coefficient to the indicator variable for the factor level with no observations. 
      \item Secondly, due to the outcome (default) being a rare event, we can have *separation*; there may be a level of a factor for which the data only has one type of event (only defaults or no defaults at all), which causes unstable estimate from ordinary Maximum Likelihood Estimation (MLE) because the optimization routine will iteratively push the estimate towards $\pm\infty$. 
  \item After choosing the specification for laon age, we will consider if we can drop any other variables. For the other variables, we have a fairly large collection of continuous and discrete predictors, so we use Lasso (L1 regularization) to decide what to include in the model. 
  \item After CV and Lasso, we will fit the un-regularized model with the chosen predictors. If there are statistically insignificant predictors in the last model, we will re-assess multicollinearity and use CV to compare a reduced model to the full model. 
\end{itemize}
   
Lastly, note that we will do all variable selection work on a training subset (chosen to be the first 80\% of time points) while setting aside a testing subset (last 20\% of time points) for final model calibration. These global subsets will be created before any CV is conducted. The train/test folds in CV for loan age specification will be subsets of the global training set.

## Set up Global Train/Test Sets
```{r, eval=FALSE}
# Load libraries for custom CV implementation
library(data.table)
library(splines)

# Create global training and holdout subsets
train_prop <- 0.8 
global_train_end <- ceiling(length(months) * train_prop)
train_mos <- months[1:global_train_end]
df_train <- full_df %>% filter(date %in% train_mos)
df_test <- full_df %>% filter(date %notin% train_mos)
```

The `glmnet` package has functions for fitting an Elastic Net GLM, but this model can be reduced into a Lasso or Ridge GLM by specifying the `alpha` parameter to be `0` or `1`.

## CV for Age of Loan Specification
```{r, eval=FALSE}
# Prepare data.table for CV
dt <- as.data.table(df_train)
dt[, date := as.IDate(date)]
dt[, y := as.integer(y)]

# Encode categorical variables in data.table as factors
cat_vars <- c("deliq_num", "occupancy_status", "region", 
              "loan_purpose")
for (v in cat_vars) dt[, (v) := as.factor((v))]


# Create random sample of loans to be used for CV
set.seed(1) # reproducibility
sample_frac <- 0.15
all_ids <- unique(dt$loan_sequence_number)
sample_ids <- sample(all_ids, size=ceiling(sample_frac*length(all_ids)), replace=FALSE)
dts <- dt[loan_sequence_number %in% sample_ids] # sampled data

cat("Sampled", length(sample_ids), "loans out of", length(all_ids), "\n")
cat("Sampled", nrow(dts), "rows out of", nrow(dt), "\n")

dts <- dt

# Create table telling us which rows correspond to each time period
date_tab <- dts[, .N, by=date][order(date)]
date_tab[, end_row := cumsum(N)]
date_tab[, start_row := shift(end_row, fill = 0) + 1]


# Function to map data into the indices defining each CV fold
make_folds <- function(date_tab, init_periods, assess_periods, step) {
  
  m <- nrow(date_tab)
  stopifnot(init_periods + assess_periods <= m)

  fold_splits <- seq(from=init_periods, to=m-assess_periods, by=step)
  
  return(
    data.table(
      fold = seq_along(fold_splits),
      train_end_row  = date_tab$end_row[fold_splits],
      test_start_row = date_tab$start_row[fold_splits + 1],
      test_end_row   = date_tab$end_row[fold_splits + assess_periods], 
      train_end_date = date_tab$date[fold_splits]
      )
  )
}

# Create CV folds
folds <- make_folds(date_tab, init_periods=30, assess_periods=8, step=8)
cat("Number of folds:", nrow(folds))



# Vector of all non-age covariate names
other_x <- setdiff(names(dts), 
                   c("loan_sequence_number", "date", "y", "loan_age"))

# Right-hand-side formula string
rhs <- paste(other_x, collapse=" + ")

# Create list of different model formulas 
specs <- list(
  age_linear = as.formula(paste0("y ~ loan_age_pct + ", rhs)),
  age_log1p = as.formula(paste0("y ~ log1p(loan_age_pct) + ", rhs)), 
  bins = as.formula(paste0("y ~ age_group_factor", rhs)), 
  age_spline = as.formula(paste0("y ~ ns(loan_age_pct, df=4) + ", rhs))
)


# Binary entropy loss function (with padding for numerical stability)
be_loss <- function(y, p, eps=1e-15){
  p <- pmin(pmax(p, eps), 1-eps)
  return(-mean(y*log(p) + (1-y)*log(1-p)))
}


# Function to evaluate a model specification on CV folds
eval_spec <- function(x, y, folds, loss_func) {
  
  n_folds <- nrow(folds)
  out_list <- vector("list", n_folds)  # store per-fold results
  
  for (i in seq_len(n_folds)) {
    
    tr_end <- folds$train_end_row[i]
    te_s <- folds$test_start_row[i]
    te_e <- folds$test_end_row[i]
    
    x_train <- x[1:tr_end, , drop = FALSE]
    x_test <- x[te_s:te_e, , drop = FALSE]
    y_train <- y[1:tr_end]
    y_test <- y[te_s:te_e]
    
    # Fit Lasso logistic regression
    fit <- glmnet(
      x = x_train,
      y = y_train,
      family = "binomial",
      alpha = 1, # alpha = 1 for Lasso
      nlambda = 50
    )
    
    # Choose the last lambda for prediction 
    lambda_use <- tail(fit$lambda, 1)
    
    p <- as.numeric(
      predict(fit, newx = x_test, type = "response", s = lambda_use)
    )
    
    loss <- loss_func(y_test, p)
    
    # one-row data.table for this fold
    out_list[[i]] <- data.table(
      fold = folds$fold[i],
      loss = loss,
      train_end_month = folds$train_end_month[i]
    )
    
    cat("\t", "Done fold ", i) # status update
  }
  
  # combine all folds
  rbindlist(out_list)
}


# Conduct CV (train and test the models on each fold)
cv_results <- rbindlist(
  lapply(names(specs), function(model) { 
    
    cat("Working on model", model) # status update
           
    formula <- specs[[model]]
    
    needed_vars <- all.vars(formula)
    dts_small <- dts[, ..needed_vars]
    
    x_mat <- model.matrix(formula, data = dts_small)[, -1, drop = FALSE]
    y_vec <- dts_small$y
  
    res <- eval_spec(x_mat, y_vec, folds, be_loss)
    res[, spec := model]
    res
  })
)


cv_summary <- cv_results[, .(
  mean_loss = mean(loss), 
  sd_loss = sd(loss), 
  se_loss = sd(loss)/sqrt(.N)
  ), 
  by = spec][order(mean_loss)]

cv_summary


best_spec <- cv_summary$spec[1]
best_formula <- specs[[best_spec]]
cat("Best specification for loan age component of model:", best_spec) 
```

*Decide on final spec to use while taking into account interpretability*

*STILL NEED TO REBOOT R THEN DO THE CV ABOVE (IT WILL TAKE A WHILE)*  

## Lasso for Large-Scale Variable Selection

We use the ability of Lasso to shrink coefficient estimates to zero to help determine what variables to keep. Since factors will be converted into a collection of indicators for each level, if Lasso ends up dropping some but not all of these indicators then we will still keep that factor in the model. 

```{r, eval=FALSE}


```
